Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-04-27 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
clarified what we lose vs. atomic lock proofs

\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the National Science Foundation via grant \#1811894.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Roshan Sharma\inst{1} \and
Shengyi Wang\inst{2}\orcidID{?} \and
Alexander Oey\inst{3} \and
Anastasiia Evdokimova\inst{4} \and
Lennart Beringer\inst{2}\orcidID{?} \and
William Mansky\inst{4}\orcidID{0000-0002-5351-895X}}

@@

\subsection{Two Styles of Lock Specification}
Soon after the introduction of concurrent separation logic~\cite{csl}, both Gotsman et al.~\cite{gotsman} and Hobor et al.~\cite{oraclesematic} extended it with support for C-style storable locks, where a lock is a special memory location designated as coordinating access to some other resource. Their lock specifications were essentially equivalent, and of the form:
$$\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}$$
Each lock $\ell$ is associated with a ``lock invariant'' $R$, an arbitrary separation logic predicate representing the protected resource, which is gained by a thread that acquires the lock and must be restored when the lock is released. At the time, there was no formal treatment of atomic operations in CSL, so actual lock implementations were beyond the scope of verification: these specs were taken as axioms, or proved through direct appeal to data-race-free guarantees of specific memory models.

Later, the TaDA logic~\cite{tada} presented an alternative set of lock specifications, based on the newly introduced notion of logical atomicity:
$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(U(\ell) * R) \vee L(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it.

@@

Our proof technique is developed in a combination of two Coq-based separation logic verification systems, the Verified Software Toolchain (VST) and Iris. VST~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs. It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run. VST directly implements the concurrent separation logic of Hobor et al.~\cite{oracle-sematics}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.

Iris~\cite{iris} is a framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''. Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state. Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity. All proofs described in this paper have been formalized in this VST+Iris setting.
%What else do we need to say?

@@

\section{Locks and Atomicity}
\label{atomic}
Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_insert}\ t\ k\ v)\}

@@

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock. Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$. This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$. Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant. The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory. This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock. They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g. the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g. P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.

@@

In a \emph{fine-grained} implementation, there are multiple locks on different pieces of the data structure, so the ghost state of each lock invariant cannot be the abstract state of the entire data structure. Instead, the ghost state represents the abstract state of the locked section alone, and the abstract state of the data structure is derived from the composition of the states of the locked components. Each locked component $c$ has a piece of ghost state $\mathsf{ghost}\_c$ that is split between the lock invariant and the top-level abstract state. At the end of a critical section for $c$, we atomically access the top-level abstract state, combine the two halves of $\mathsf{ghost}\_c$, update $\mathsf{ghost}\_c$ to reflect any local changes, and then show either that the update to $c$ does not change the top-level state of the data structure, or that the update to $c$ changes the top-level state to one that satisfies the atomic postcondition for the function. For instance, when deleting a node from a binary search tree, we rotate nodes to reconstruct a valid tree; at the end of the critical section for each rotation, we show either that the values in the tree are unchanged (and the tree is still well-ordered), or that we have removed the node to be deleted. %elaborate on two-level structure

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. %outline how this worked for them

In particular, in the fine-grained case, there is no need to create extra ghost state to abstract the state of each locked component, and then synthesize it into a top-level abstract state for the data structure. Instead, each lock acquire and release can atomically access the top-level abstract state, and directly make changes there. It is even possible for a thread to acquire a lock without owning a share of the lock assertion: the lock assertion can be extracted directly from the top-level state during the atomic \texttt{acquire} operation, and returned when \texttt{acquire} returns. This completely circumvents the recursive-lock reasoning needed for hand-over-hand locking with lock-invariant specs. %This comes later in the current version.

@@

@@

\section{Conclusion}
\label{conclusion}
While TaDA-style atomic specs for locks have been widely adopted%?
, the implications of the switch away from invariant-based specs had not been thoroughly examined. We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires significantly more complex proofs than using atomic lock specs (especially for fine-grained implementations). The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the two-level proof structure we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures. In all other settings, however, we should strongly prefer to use atomic specifications for locks to simplify our proofs. Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof may be an enormous effort, we believe we have shown it to be worth the trouble in the long run.

>>>>

\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the National Science Foundation via grant \#1811894.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Roshan Sharma\inst{1} \and
Shengyi Wang\inst{2}\orcidID{?} \and
Alexander Oey\inst{3} \and
Anastasiia Evdokimova\inst{4} \and
Lennart Beringer\inst{2}\orcidID{?} \and
William Mansky\inst{4}\orcidID{0000-0002-5351-895X}} %confirm author order

@@

\subsection{Two Styles of Lock Specification}
Soon after the introduction of concurrent separation logic~\cite{csl}, both Gotsman et al.~\cite{gotsman} and Hobor et al.~\cite{oraclesematic} extended it with support for C-style storable locks, where a lock is a special memory location designated as coordinating access to some other resource. Their lock specifications were essentially equivalent, and of the form:
$$\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}$$
Each lock $\ell$ is associated with a ``lock invariant'' $R$, an arbitrary separation logic predicate representing the protected resource, which is gained by a thread that acquires the lock and must be restored when the lock is released. At the time, there was no formal treatment of atomic operations in CSL, so actual lock implementations were beyond the scope of verification: these specs were taken as axioms, or proved through direct appeal to data-race-free guarantees of specific memory models. In garbage-collected languages that do not deallocate locks, the assertion $\ell \islock R$ can be freely duplicated and shared among threads accessing $R$; if we want to track ownership and eventually deallocate the lock, we can instead split the assertion into fractional shares such that $\ell \islock_{\pi_1} R * \ell \islock_{\pi_2} R = \ell \islock_{\pi_1 + \pi_2} R$, and then allow deallocation only when we own the full share $\ell \islock_1 R$.

Later, the TaDA logic~\cite{tada} presented an alternative set of lock specifications, based on the newly introduced notion of logical atomicity:
$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

Our proof technique is developed in a combination of two Coq-based separation logic verification systems, the Verified Software Toolchain (VST) and Iris. VST~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs. It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run. VST directly implements the concurrent separation logic of Hobor et al.~\cite{oraclesematic}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.

Iris~\cite{iris} is a framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''. Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state. Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity. Iris has also been used as the basis for proofs with atomic lock specifications, such as the work of Krishna et al.~\cite{templates}. All proofs described in this paper have been formalized in this VST+Iris setting.
%What else do we need to say?

@@

\section{Locks and Atomicity}
\label{atomic}
In this section, we outline our approach to connecting lock invariants and atomic specifications, and contrast it with the atomic-lock-based approach.

Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_insert}\ t\ k\ v)\}

@@

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock. Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$. This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$. %clarify share
Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant. The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory. This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock. They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g. the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g. P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.

@@

In a \emph{fine-grained} implementation, there are multiple locks on different pieces of the data structure, so the ghost state of each lock invariant cannot be the abstract state of the entire data structure. Instead, the ghost state represents the abstract state of the locked section alone, and the abstract state of the data structure is derived from the composition of the states of the locked components. Each locked component $c$ has a piece of ghost state $\mathsf{ghost}\_c$ that is split between the lock invariant and the top-level abstract state. At the end of a critical section for $c$, we atomically access the top-level abstract state, combine the two halves of $\mathsf{ghost}\_c$, update $\mathsf{ghost}\_c$ to reflect any local changes, and then show either that the update to $c$ does not change the top-level state of the data structure, or that the update to $c$ changes the top-level state to one that satisfies the atomic postcondition for the function. For instance, when deleting a node from a binary search tree, we rotate nodes to reconstruct a valid tree; at the end of the critical section for each rotation, we show either that the values in the tree are unchanged (and the tree is still well-ordered), or that we have removed the node to be deleted. %elaborate on two-level structure
As we will see in section~\ref{hand-over-hand}, the lock invariant for each component must also carefully account for the ownership of both that component's lock and the locks of related nodes (e.g., child nodes in a tree).

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

Note that in this case, the private precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node. A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state. More generally, in this approach there is no need to track ownership of the locks of individual nodes, a source of considerable complexity in the lock-invariant-based specs. %This comes later in the current version.

@@

This pattern is the main source of complexity in our proofs that is not present in proofs with atomic lock specifications. For instance, in the hand-over-hand locking proofs of Krishna et al.~\cite{templates}, the only information a node holds about its children is the fact that they are in the data structure; their lock assertions are held in the global abstract state along with the rest of their resources. This significantly simplifies the specification and proof of the hand-over-hand synchronization.

@@

\section{Conclusion}
\label{conclusion}
While TaDA-style atomic specs for locks have been widely adopted%?
, the implications of the switch away from invariant-based specs had not been thoroughly examined. We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations). The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures. When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs. Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof will be a considerable effort, using the newer style of lock specification will simplify all future proofs about lock-based data structures.

### 2022-04-27 concurrency/ASL 2022/intro.tex
clarified what we lose vs. atomic lock proofs

In this paper, we adapt the latter approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t. the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs. We demonstrate that 1) \textbf{this approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs are much more complicated than those built with style 2 lock specs}. The key difference is that in style 1 we must associate each locked component with some resources and then relate those resources to the state of the data structure as a whole, while in style 2 each lock access can interact directly with the full state of the data structure. However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise. To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique may be useful in tools like VST where the older style is deeply integrated into the tool.

>>>>

In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t. the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs. We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}. The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access. However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise. To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique may be useful in tools like VST where the older style is deeply integrated into the tool.
